Effects of bariatric surgery and weight loss on resting state functional connectivity

### 2021-02-05 manuscript/draft_rsfmri.Rmd
added supplementary tables

---
title: "Effects of bariatric surgery and weight loss on resting state functional connectivity"
author:
- Hannah Sophie Heinrichs* [1]
- Frauke Beyer* [1][2]
- Kristin Prehn
- Jürgen Ordemann
- Agnes Floel
- A. Veronica Witte [1][2][3]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3
  Day Clinic for Cognitive Neurology, University Clinic Leipzig
output:
  bookdown::pdf_book:
    extra_dependencies: "subfig"
    fig_caption: yes
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
    extra_dependencies: "subfig"
    fig_caption: yes
    number_sections: yes
bibliography: "bibliography.bib"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

```{r "To do", include=FALSE}
# [ ] Upload auf Neurovolt, cf. cobidas reporting guidelines
# [x] Swap mFD with logmFD ??
```

@@

\newpage
# Introduction
Obesity is a worldwide health issue, enchaining huge personal and societal costs. Excess amount of fat not only affects cardiovascular and metabolic health, but also increases the risk for cognitive decline and dementia later in life.
Conservative treatment options including dietary modification, physical activity and behavioral therapy often do not yield the desired weight loss, especially in patients with very high BMI (> 35 kg/m2). Here, bariatric surgery is a viable option to rapidly induce weight loss and improve glycemic status.
Roux-en-Y gastric bypass (RYGB) is the most effective surgical procedure [@Berthoud_2011]. Here, a small pouch is formed from the proximal stomach and connected to the jejunum, disconnecting the larger part of the stomach and the proximal part of the small intestine intact from the digestive flow. Other techniques like vertical sleeve gastrectomy (VSG) and gastric banding (GB) reduce stomach volume, but leave the small intestine unchanged [@Mulla_2017].
RYGB-induced weight loss is in part mediated by malabsorption of certain nutrients, especially fat, and reduced digestion efficiency [@Berthoud_2011] while the role of increased energy expenditure after RYGB is still unclear [@Yarmush_2017;@Lamarca_2019]. Most importantly, RYGB alters the signaling cascades between the gut, fat tissue and the brain and thereby induces profound changes in the regulation of food intake and energy homeostasis [@Berthoud_2011]. RYBG-patients frequently report altered smell and taste perception, i.e. increased sensitivity to sweet taste, show less desire to consume high-calorie and high-fat foods and actually reduce intake of such food items [@Mulla_2017, @Berthoud_2011].  These changes in appetitive and reward signaling likely relate to altered gut-brain communication, mediated by afferent vagal nerve signaling and gut hormones like GLP-1, PYY and ghrelin [ @Brutman_2019].

Yet, precise mechanisms how bariatric surgery leads to altered appetitive and reward signaling are yet to be elucidated.
Resting state functional magnetic resonance imaging (rsfMRI) is a powerful technique to investigate the dynamic organization of the brain via the correlation of spontaneous neural activity. Anatomically distinct brain regions show functional connectivity (FC), i.e. correlated neural activity over time, in the absence of a cognitive task. Crucially, these brain networks correspond to networks activated during specific cognitive processes, e.g. visual, auditory and somatosensory processing, or higher-order cognitive functions, i.e. attention and executive control [@Smith_2009], and can thus be seen as default configuration of the brain. In the context of bariatric surgery, the default mode network, a higher-order network, involved in interoception and governing shifts between external-internal processes, and the reward network, processing hedonic value and internal motivation, are promising candidates to mediate altered gut-brain signaling.
The default mode network has been discovered as a “task-negative” network, and is involved in a variety of largely internally-directed cognitive functions including interoception, mind wandering and memory. Its core hubs are the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (mPFC) and the inferior and lateral parietal cortex (approx. BA 39) [@Raichle_2015].
Higher BMI and obesity have been associated with differences in DMN FC in several resting state and task-based fMRI studies [@Beyer_2017; @Kullmann_2011; @Chao_2018; @Wijngaarden_2015, @Borowitz_2020; @Ding_2020; @Doucet_2017; @Sadler_2018].
Despite the large heterogenity and low median power, these findings might be summarized as a pattern of decreased within DMN FC and increased FC of DMN regions to other networks, i.e. salience and sensory networks. This has been interpreted as impaired interoceptive processing failing to adequately balance internal and external behavioral cues.
Longitudinal studies largely support this view. Increased activation of the DMN in a food-viewing paradigm, i.e. insufficient down-regulation of DMN, has been shown to correlate with appetite measures after an overnight fast [@Tregellas_2011, @Wijngaarden_2015], an effect that could be partially reversed by an exercise intervention [@McFadden_2013]. In a bariatric surgery study, [@Li_2018] reported a reduction of local FC in mid-line DMN regions in surgical patients compared to an obese control group. Simultaneously, PCC and medial prefrontal cortex were more strongly connected to dorsolateral prefrontal cortex, a region implicated in cognitive control. Another study in bariatric surgery patients showed that heightened connectivity of the PCC with anterior superior temporal gyrus (STG) before the surgery decreased over the course of one year [@Olivo_2017]. Similarly, the profile of DMN connectivity with frontal regions, including ACC, frontal superior gyrus and the right OFC,  was similar in patients after bariatric surgery and normal weight controls, but dissimilar to that of obese controls [@Frank_2013].

Another important brain network in the context of obesity is the reward network, comprising the ventromedial prefrontal cortex (vmPFC), the striatum (STR), the amygdala (AMY) and the anterior insula (antINS) [@Odoherty_2004;@Liu_2011]. These brain regions have been found to guide food valuation processes and decision-making in humans [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018]. Frequently, obesity has been associated with hyperactivation of reward network regions during anticipation of (high-caloric) food cues, and in contrast, reduced reward region activation to actual taste of these foods [@Devoto_2018; @Garcia_2014; @Meng_2020; @Stoeckel_2009; though see @Morys_2020;]. Food cue hyperactivation has been identified as a main neural vulnerability factor predicting future weight gain [@Stice_2015]. Resting state fMRI studies have partly mirrored these results, showing increased local FC of reward network regions, i.e. Nacc, vmPFC, putamen, insula [@Contreras_2017; @Coveleskie_2015;@Hogenkamp_2016], and altered connectivity with salience, homeostatic and sensorimotor networks [@Lips_2014;  @Wijngaarden_2015]. Imaging studies in bariatric surgery patients provided first evidence that hyperconnectivity between reward network regions (e.g. putamen, OFC) might be normalized to some extent by the surgery [@Duan_2020; @Wiemerslage_2016]. Possible, the reconfiguration of the fronto-striatal brain networks might be related to altered gut signaling , e.g. changes in ghrelin levels, via hypothalamic-striatal projections [@Li_2019; @Karra_2013, but @Zoon_2018] .
Thus, while there is some evidence hinting to a role for DMN and reward network connectivity in altered gut-brain communication after bariatric surgery, the existing evidence is inconclusive. Most studies have investigated small cohort of patients, often without adequate control group, and employed exploratory statistics [@George_2016].

@@

\newpage
# Methods

```{r "Load data", include=FALSE}
#final <- create_sample_df(group = "both", tp = "all")
final <- read.csv(file = "../report/final.csv")

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study investigated the effects of bariatric surgery on brain structure and function in a prospective design at the Charité Berlin. For more details see @Prehn_2020. We used all data acquired until April 2019.  
The study protocol was in accordance with the Helsinki Declaration and approved by the local ethics committee. The study was registered at clinicaltrials.gov as NCT01554228. We preregistered the present resting state analyses on the [OSF](https://osf.io/yp42s). We made [additional changes](https://osf.io/59bh7/) to the preregistration after preprocessing as we realized some aspects of the analysis were inadequately described in the first preregistration.   
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA -->
Participants were recruited from the Center for Bariatric and Metabolic Surgery. Inclusion criteria were, in accordance with BARS guidelines, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013].
Participants were aged between $18$ and $70$ and had no history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score $< 34$).
In total, $51$ participants out of the originally enrolled $69$ subjects received MRI, and we analyzed a total of $101$ data points. Five data points of three subjects had to be excluded due to bad anatomical image quality (see section \@ref{sec:Anatomical Quality control}).
<!-- DESCRIPTIVE STATISTICS -->
The final sample entailed `r nrow(subj_df)` morbidly obese individuals (`r nrow(filter(subj_df,Sex == "f"))` females; aged `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`). Participants either underwent surgery (n = `r cs["Sum","IG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female), who waited for their health insurance's approval to undergo surgery. Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery or after the baseline appointment. Analyses were performed on all participants that provided at least one data point of fMRI data (detailed information on data availabile. In more detail, `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for only one time point. In the intervention group, data is not missing completely at random, as the pre-surgery timepoint MRI could not be obtained in some patients due to the limited scanner bore.
```{r tableBaselineCharacteristics, include=FALSE}
#dfBL <- create_sample_df(group = "all", tp = "BL")
dfBL <- readRDS(file = "../report/dfBL.rds")

@@

<!-- STUDY PROCEDURE --> Subjects arrived in the morning (between 07:00 and 12:00) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one our break for <!-- standadized??  -->breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020).

```{r figBMIdescr, fig.align='center', out.width="100%", fig.cap="Trajectoriy of BMI separately for the bariatric surgery group (BARS) and the waiting-list control group (NBARS); individual trajectories are plotted in transparent, mean trajectories including standard deviations in opaque colors.", echo=FALSE}
# eval = TRUE, fig.asp = 0.7, fig.width = 6
readRDS(file = "../report/fig/fig_BMIdescr.rds")
```
<!-- BASELINE CHARACTERISTICS --> Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`). Change in BMI throughout the study is depicted in \@ref(fig:figBMIdescr).
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).
<!-- STUDY PROCEDURE --> Subjects arrived in the morning (between 07:00 and 12:00) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropocentric measurements before having a one our break for <!-- standadized??  -->breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020).

@@

#### Functional Quality control
Resting state fMRI data quality control was performed according to the protocol by @Ciric_2018.
As the expected average head motion was high in this sample, it was difficult to determine an appropriate exclusion threshold which would not lead to the exclusion of a high number of individuals [@Beyer_2020]. Thus, we did not exclude any participants based on mean FD estimates beyond those who failed the anatomical preprocessing. We applied best-practice motion correction (ICA-AROMA and CC, plus optionally GSR) and monitored the reduction of motion-related artifact according to current recommendations [@Ciric_2018]. In addition, for models showing a significant time-by-group interaction, we performed a sensitivity analysis excluding the 10% data sets with highest mean FD.
```{r carpetplot, out.width="50%", out.height="30%", fig.cap="Carpet plot for quality control", fig.subcap = c("Example 1 with strong amount of structured noise", "Example 2 with less amount of structured noise"), fig.show='hold',fig.align='center', echo=FALSE}
carpets=c("../report/carpetplotadi006fu2.jpeg","../report/carpetplotadi16fu.jpeg")
knitr::include_graphics(carpets)
```
On the single-subject level, we used `plot_carpet` from `niworkflows.viz.plots` to generate carpet plots which depict denoising quality. Figure \@ref(fig:carpetplot) shows an exemplary carpet plot with (from top to bottom): percent outliers defined by AFNI, mFD, `OutlierCount`, DVARS (spatial root mean square of the data after temporal differencing),voxelwise timeseries from minimally preprocessed, AROMA non-aggressive and aggressive denoising, CC and CC + GSR data (red: GM voxels, green: WM voxels, blue: ventricle, green: cerebellum). All carpet plots were visually checked and if there was still structured noise in the CC+GSR-denoised functional data (such as black/white stripes, signal dropout), the participant was given a rating of 1 in `QA residuals`.
We evaluated mFD-FC relationships by computing a functional connectome based on [@Power_2012]. In more detail, we used the MNI coordinates of 246 cortical and subcortical spherical seed regions with 5mm radius in `NiftiSpheresMasker`, and calculated the Spearman rank correlation between these time series. Then, we correlated mean FD and QC values for each node across participants, and plotted the resulting mFD-FC correlation values in histogram (see Figure \@ref{exampleFC}. We reported the mFD-FC correlation per preprocessing scheme.
Further, we evaluated the distance-dependence of mFD-FC correlations by correlating the euclidean distance between nodes with the mFD-QC correlation and report mean, median and standard deviations of mFD, DVARS and number of outliers over all participants.

### Aggregated FC values
To extract aggregated FC values, we first calculated the mean DMN and reward network over all participants and timepoints, adjusted for age and sex. We used GSR regressed data as input and clusterwise bootstrapping with N=999. Network masks were formed from all voxels within clusters which survived a clusterwise multiple comparison correction of FWE-corrected $p < 0.05$. We extracted the average connectivity from these masks for GSR regressed data.
*In the preregistration we wrote: "We will use the average of connectivity values within masks of the respective networks as aggregate network connectivity measure. The mask will include all voxels surviving FWE-correction of p<0.05 in a one-sample t-test of the individual connectivity maps across all participants and time points, calculated with the modified SwE option in the SwE toolbox"*

@@

We did not adjust for baseline BMI as described in the preregistration, but explored the relationship of average and change in BMI in a separate group of models.
As proposed by @Guillaume_2014, we calculated the between- and within subject centered values of BMI. This model, containing average BMI and BMI variability, allowed us to disentangle the differential effects of overall BMI and change in BMI on FC. We first estimated their effects in a model adjusting for baseline age and sex (Model 2a) and then additionally controlling for log mean FD (Model 2b).
As we previously reported correlated change in BMI and mFD in this sample [@Beyer_2019], we explored a refined model including average BMI and log mean FD and variability in both measures, along with baseline age and sex (Model 3). Here, we aimed to see whether any effect of change in BMI would be detectable when adjusting for the change in head motion.
<!-- POST HOC -> können wir nochmal überlegen, ob wir das drinhaben wollen. -->For a better understanding of the unique contribution of average and longitudinal change FD measures, we separately employed a third post-hoc, exploratory model: $FC = between-subject FD + within-subject FD$ with nuisance covariates age and sex.

@@

\newpage
# Results

## Quality control
We performed different quality controls for the rsfMRI data.
First, we inspected parameters of head motion (mFD) and a measure of whole-brain signal variation which is reflective of noise in the data (DVARS).
As described in @Beyer_2020, there was a group-by-time interaction on mFD. The intervention group reduced in head motion in the follow-up assessments compared to the control group.
Further, DVARS did not qualitatively differ between groups and time points, and the correlation of DVARS and mFD remained similar over time points.
We concluded that even though there was a decrease in head motion in the intervention versus control group, the impact of head motion on imaging data remained similar over the study period in both groups.

>>>>

---
title: "Effects of bariatric surgery and weight loss on resting state functional connectivity"
author:
- Hannah Sophie Heinrichs* [1]
- Frauke Beyer* [1][2]
- Evelyn Medawar [1]
- Kristin Prehn [3][4]
- Jürgen Ordemann [8][9]
- Agnes Floel [5][6][7]
- A. Veronica Witte [1][2][10]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3 Department of Neurology and NeuroCure Clinical Research Center, Charité University Medicine, Berlin \newline 4 Department of Psychology, Medical School Hamburg, Hamburg \newline 5 Department of Neurology, University of Greifswald, Greifswald \newline 6 German Center for Neurodegenerative Diseases, Standort Rostock/Greifswald, Greifswald \newline 7 Center for Stroke Research, Charité University Medicine, Berlin \newline 8 Center for Bariatric and Metabolic Surgery, Charité University Medicine, Berlin \newline 9 Zentrum für Adipositas und Metabolische Chirurgie, Vivantes Klinikum Spandau, Berlin \newline 10 Day Clinic for Cognitive Neurology, University Clinic Leipzig 
output:
  bookdown::pdf_book:
    extra_dependencies: "subfig"
    fig_caption: yes
    number_sections: yes
    toc: yes
bibliography: "bibliography.bib"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
- \usepackage{setspace}
- \onehalfspacing
indent: true
---

```{r "To do", include=FALSE}
# [x] Upload auf Neurovolt, cf. cobidas reporting guidelines
# [x] Swap mFD with logmFD ??
```

@@

\newpage
# Introduction
Obesity is a worldwide health issue, enchaining huge personal and societal costs. Excess amount of fat not only affects cardiovascular and metabolic health, but also increases the risk for cognitive decline and dementia later in life.
Conservative treatment options including dietary modification, physical activity and behavioral therapy often do not yield the desired weight loss, especially in patients with very high BMI (> 35 kg/m2). Here, bariatric surgery is a viable option to rapidly induce weight loss and improve glycemic status.
Roux-en-Y gastric bypass (RYGB) is the most effective surgical procedure [@Berthoud_2011]. Here, a small pouch is formed from the proximal stomach and connected to the jejunum, disconnecting the larger part of the stomach and the proximal part of the small intestine intact from the digestive flow. Other techniques like vertical sleeve gastrectomy (VSG) and gastric banding (GB) reduce stomach volume, but leave the small intestine unchanged [@Mulla_2017].
RYGB-induced weight loss is in part mediated by malabsorption of certain nutrients, especially fat, and reduced digestion efficiency [@Berthoud_2011] while the role of increased energy expenditure after RYGB is still unclear [@Yarmush_2017;@Lamarca_2019]. Most importantly, RYGB alters the signaling cascades between the gut, fat tissue and the brain and thereby induces profound changes in the regulation of food intake and energy homeostasis [@Berthoud_2011]. RYBG-patients frequently report altered smell and taste perception, i.e. increased sensitivity to sweet taste, show less desire to consume high-calorie and high-fat foods and actually reduce intake of such food items [@Mulla_2017; @Berthoud_2011].  These changes in appetitive and reward signaling likely relate to altered gut-brain communication, mediated by afferent vagal nerve signaling and gut hormones like GLP-1, PYY and ghrelin [@Brutman_2019].

Yet, precise mechanisms how bariatric surgery leads to altered appetitive and reward signaling are yet to be elucidated.
Resting state functional magnetic resonance imaging (rsfMRI) is a powerful technique to investigate the dynamic organization of the brain via the correlation of spontaneous neural activity. Anatomically distinct brain regions show functional connectivity (FC), i.e. correlated neural activity over time, in the absence of a cognitive task. Crucially, these brain networks correspond to networks activated during specific cognitive processes, e.g. visual, auditory and somatosensory processing, or higher-order cognitive functions, i.e. attention and executive control [@Smith_2009], and can thus be seen as default configuration of the brain. In the context of bariatric surgery, the default mode network, a higher-order network, involved in interoception and governing shifts between external-internal processes, and the reward network, processing hedonic value and internal motivation, are promising candidates to mediate altered gut-brain signaling.
The default mode network has been discovered as a “task-negative” network, and is involved in a variety of largely internally-directed cognitive functions including interoception, mind wandering and memory. Its core hubs are the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (mPFC) and the inferior and lateral parietal cortex (approx. BA 39) [@Raichle_2015].
Higher BMI and obesity have been associated with differences in DMN FC in several resting state and task-based fMRI studies [@Beyer_2017; @Kullmann_2011; @Chao_2018; @Wijngaarden_2015; @Borowitz_2020; @Ding_2020; @Doucet_2017; @Sadler_2018].
Despite the large heterogenity and low median power, these findings might be summarized as a pattern of decreased within DMN FC and increased FC of DMN regions to other networks, i.e. salience and sensory networks. This has been interpreted as impaired interoceptive processing failing to adequately balance internal and external behavioral cues.
Longitudinal studies largely support this view. Increased activation of the DMN in a food-viewing paradigm, i.e. insufficient down-regulation of DMN, has been shown to correlate with appetite measures after an overnight fast [@Tregellas_2011; @Wijngaarden_2015], an effect that could be partially reversed by an exercise intervention [@McFadden_2013]. In a bariatric surgery study, [@Li_2018] reported a reduction of local FC in mid-line DMN regions in surgical patients compared to an obese control group. Simultaneously, PCC and medial prefrontal cortex were more strongly connected to dorsolateral prefrontal cortex, a region implicated in cognitive control. Another study in bariatric surgery patients showed that heightened connectivity of the PCC with anterior superior temporal gyrus (STG) before the surgery decreased over the course of one year [@Olivo_2017]. Similarly, the profile of DMN connectivity with frontal regions, including ACC, frontal superior gyrus and the right OFC,  was similar in patients after bariatric surgery and normal weight controls, but dissimilar to that of obese controls [@Frank_2013].

Another important brain network in the context of obesity is the reward network, comprising the ventromedial prefrontal cortex (vmPFC), the striatum (STR), the amygdala (AMY) and the anterior insula (antINS) [@Odoherty_2004;@Liu_2011]. These brain regions have been found to guide food valuation processes and decision-making in humans [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018]. Frequently, obesity has been associated with hyperactivation of reward network regions during anticipation of (high-caloric) food cues, and in contrast, reduced reward region activation to actual taste of these foods [@Devoto_2018; @Garcia_2014; @Meng_2020; @Stoeckel_2009; though see @Morys_2020]. Food cue hyperactivation has been identified as a main neural vulnerability factor predicting future weight gain [@Stice_2015]. Resting state fMRI studies have partly mirrored these results, showing increased local FC of reward network regions, i.e. Nacc, vmPFC, putamen, insula [@Contreras_2017; @Coveleskie_2015;@Hogenkamp_2016], and altered connectivity with salience, homeostatic and sensorimotor networks [@Lips_2014;  @Wijngaarden_2015]. Imaging studies in bariatric surgery patients provided first evidence that hyperconnectivity between reward network regions (e.g. putamen, OFC) might be normalized to some extent by the surgery [@Duan_2020; @Wiemerslage_2016]. Possible, the reconfiguration of the fronto-striatal brain networks might be related to altered gut signaling , e.g. changes in ghrelin levels, via hypothalamic-striatal projections [@Li_2019; @Karra_2013; but @Zoon_2018] .
Thus, while there is some evidence hinting to a role for DMN and reward network connectivity in altered gut-brain communication after bariatric surgery, the existing evidence is inconclusive. Most studies have investigated small cohort of patients, often without adequate control group, and employed exploratory statistics [@George_2016].

@@

# Methods
```{r "Load data", include=FALSE}
#final <- create_sample_df(group = "both", tp = "all")
final <- read.csv(file = "../report/final.csv")

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study investigated the effects of bariatric surgery on brain structure and function in a prospective design at the Charité Berlin. For more details see @Prehn_2020. We used all data acquired until April 2019.  
The study protocol was in accordance with the Helsinki Declaration and approved by the local ethics committee. The study was registered at clinicaltrials.gov as NCT01554228. We preregistered the present resting state analyses on the [OSF](https://osf.io/yp42s). We made [additional changes](https://osf.io/59bh7/) to the preregistration after preprocessing as we realized some aspects of the analysis were inadequately described in the first preregistration.   
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA -->
Participants were recruited from the Center for Bariatric and Metabolic Surgery. Inclusion criteria were, in accordance with BARS guidelines, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013].
Participants were aged between $18$ and $70$ and had no history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score $< 34$).
In total, $51$ participants out of the originally enrolled $69$ subjects received MRI, and we analyzed a total of $101$ data points. Five data points of three subjects had to be excluded due to bad anatomical image quality (see section \@ref{sec:Anatomical Quality control}).
<!-- DESCRIPTIVE STATISTICS -->
The final sample entailed `r nrow(subj_df)` morbidly obese individuals (`r nrow(filter(subj_df,Sex == "f"))` females; aged `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`). Participants either underwent surgery (n = `r cs["Sum","IG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female), who waited for their health insurance's approval to undergo surgery. Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery or after the baseline appointment. Analyses were performed on all participants that provided at least one data point of fMRI data (detailed information on data available. In more detail, `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for only one time point. In the intervention group, data is not missing completely at random, as the pre-surgery timepoint MRI could not be obtained in some patients due to the limited scanner bore.
```{r tableBaselineCharacteristics, include=FALSE}
#dfBL <- create_sample_df(group = "all", tp = "BL")
dfBL <- readRDS(file = "../report/dfBL.rds")

@@

<!-- STUDY PROCEDURE --> Subjects arrived in the morning (between 07:00 and 12:00) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one our break for <!-- standadized??  -->breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020).
```{r figBMIdescr, fig.align='center', out.width="100%", fig.cap="Trajectoriy of BMI separately for the bariatric surgery group (BARS) and the waiting-list control group (NBARS); individual trajectories are plotted in transparent, mean trajectories including standard deviations in opaque colors.", echo=FALSE}
# eval = TRUE, fig.asp = 0.7, fig.width = 6
readRDS(file = "../report/fig/fig_BMIdescr.rds")
```
<!-- BASELINE CHARACTERISTICS --> Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`). Change in BMI throughout the study is depicted in \@ref(fig:figBMIdescr).
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).
<!-- STUDY PROCEDURE --> Subjects arrived in the morning (between 07:00 and 12:00) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropocentric measurements before having a one our break for <!-- standadized??  -->breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020).

@@

#### Functional Quality control
Resting state fMRI data quality control was performed according to the protocol by @Ciric_2018.
As the expected average head motion was high in this sample, it was difficult to determine an appropriate exclusion threshold which would not lead to the exclusion of a high number of individuals [@Beyer_2020]. Thus, we did not exclude any participants based on mean FD estimates beyond those who failed the anatomical preprocessing. We applied best-practice motion correction (ICA-AROMA and CC, plus optionally GSR) and monitored the reduction of motion-related artifact according to current recommendations [@Ciric_2018]. In addition, for models showing a significant time-by-group interaction, we performed a sensitivity analysis excluding the 10% data sets with highest mean FD.
```{r carpetplot, out.width="50%", out.height="30%", fig.cap="Carpet plot for quality control", fig.subcap = c("Example 1 with strong amount of structured noise", "Example 2 with less amount of structured noise"), fig.show='hold',fig.align='center', echo=FALSE}
carpets=c("../report/carpetplotadi006fu2.jpeg","../report/carpetplotadi16fu.jpeg")
knitr::include_graphics(carpets)
```
On the single-subject level, we used `plot_carpet` from `niworkflows.viz.plots` to generate carpet plots which depict denoising quality. Figure \@ref(fig:carpetplot) shows an exemplary carpet plot with (from top to bottom): percent outliers defined by AFNI, mFD, `OutlierCount`, DVARS (spatial root mean square of the data after temporal differencing),voxelwise timeseries from minimally preprocessed, AROMA non-aggressive and aggressive denoising, CC and CC + GSR data (red: GM voxels, green: WM voxels, blue: ventricle, green: cerebellum). All carpet plots were visually checked and if there was still structured noise in the CC+GSR-denoised functional data (such as black/white stripes, signal dropout), the participant was given a rating of 1 in `QA residuals`.
We evaluated mFD-FC relationships by computing a functional connectome based on [@Power_2012]. In more detail, we used the MNI coordinates of 246 cortical and subcortical spherical seed regions with 5mm radius in `NiftiSpheresMasker`, and calculated the Spearman rank correlation between these time series. Then, we correlated mean FD and QC values for each node across participants, and plotted the resulting mFD-FC correlation values in histograms (see supplements). We reported the mFD-FC correlation per preprocessing scheme.
Further, we evaluated the distance-dependence of mFD-FC correlations by correlating the euclidean distance between nodes with the mFD-QC correlation and report mean, median and standard deviations of mFD, DVARS and number of outliers over all participants.

### Aggregated FC values
To extract aggregated FC values, we first calculated the mean DMN and reward network over all participants and timepoints, adjusted for age and sex. We used GSR regressed data as input and clusterwise bootstrapping with N=999. Network masks were formed from all voxels within clusters which survived a clusterwise multiple comparison correction of FWE-corrected $p < 0.05$. We extracted the average connectivity from these masks for GSR regressed data.
*In the preregistration we wrote: "We will use the average of connectivity values within masks of the respective networks as aggregate network connectivity measure. The mask will include all voxels surviving FWE-correction of .05 in a one-sample t-test of the individual connectivity maps across all participants and time points, calculated with the modified SwE option in the SwE toolbox"*

@@

We did not adjust for baseline BMI as described in the preregistration, but explored the relationship of average and change in BMI in a separate group of models.
As proposed by @Guillaume_2014, we calculated the between- and within subject centered values of BMI. This model, containing average BMI and BMI variability, allowed us to disentangle the differential effects of overall BMI and change in BMI on FC. We first estimated their effects in a model adjusting for baseline age and sex (Model 2a) and then additionally controlling for log mean FD (Model 2b).
As we previously reported correlated change in BMI and mFD in this sample [@Beyer_2019], we explored a refined model including average BMI and log mFD and change in both measures, along with baseline age and sex (Model 2c). Here, we aimed to see whether any effect of change in BMI would be detectable when adjusting for the change in head motion.
<!-- POST HOC -> können wir nochmal überlegen, ob wir das drinhaben wollen. -->For a better understanding of the unique contribution of average and longitudinal change FD measures, we separately employed a third post-hoc, exploratory model: $FC = between-subject FD + within-subject FD$ with nuisance covariates age and sex.

@@

# Results

## Quality control

We performed different quality controls for the rsfMRI data.
First, we inspected parameters of head motion (mFD) and a measure of whole-brain signal variation which is reflective of noise in the data (DVARS).
As described in @Beyer_2020, there was a group-by-time interaction on mFD. The intervention group reduced in head motion in the follow-up assessments compared to the control group.
Further, DVARS did not qualitatively differ between groups and time points, and the correlation of DVARS and mFD remained similar over time points.
We concluded that even though there was a decrease in head motion in the intervention versus control group, the impact of head motion on imaging data remained similar over the study period in both groups.
```{r mFDFC, echo=FALSE}
res <- read.csv("../report/FDFCcorrelations.csv")
rownames(res) <- c("minimally processed", "AROMA", "AROMA + CC", "AROMA + CC + GSR")
colnames(res) <- c('mean FD-QC', 'median FD-QC', 'sig. vertex', 'sig. vertex BH', 'distance-FD-QC', 'pval')
```
Further, we evaluated the impact of our denoising pipelines on different quality metrics following [@Ciric_2018]. Table \@ref(tab:tableQcfc) shows that in the minimally preprocessed data, mFD and FC between nodes was significantly positively correlated (mean Spearman's r = `r round(res[1,1],2)`). This relationship was negatively associated with distance between nodes (mean Spearman's r = `r round(res[1,5],2)`).  
Denoising procedures (AROMA, AROMA+CC, AROMA+CC+GSR) gradually reduced the correlation of mFD and connectivity, and reduced the distance dependency. Contrary to previous reports, GSR did not exercabate mFD-FC distance dependency [@Power_2015]. Overall, AROMA+CC+GSR performed best, with almost no mFD-FC correlation (mean Spearman's r = `r round(res[4,1],2)`), and minimal positive distance dependence (mean Spearman's r = `r round(res[4,5],2)`). Yet, AROMA+CC also performed well, and we thus calculated our analysis models for both preprocessing schemes.
```{r tableQcfc, echo=FALSE}
tab_Qcfc <-
  knitr::kable(
    round(res, 3),
    col.names = colnames(res),
    row.names = TRUE,
    format = "latex",
    booktabs = T,
    linesep = "",
    caption = "Summary of quality metrics for different denoising pipelines across conditions and time points"
  ) %>%
  column_spec(c(2:6), width = "1.8cm")
tab_Qcfc
```
